Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2016-12-29 pivot/paper.tex
More work on draft

\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage{lipsum}
\usepackage{semantic}

@@

@@

We unify this via dot

Could this be done without type provider?

@@

  ○ We use simple nominal typing
\subsection{Column names}
\label{sec:analysis-col}

@@

  ○ We use type providers to provide available options
  (e.g. something)
\subsection{Dictionary syntax}
\label{sec:analysis-and}
  ○ We create the "and blah. and blah. then" pattern
  (e.g. grouping and sorting)

@@

>>>>

\usepackage{microtype}%if unwanted, comment out or use option "draft"
\usepackage{pifont}
\usepackage{fp}
\usepackage{graphicx}
\usepackage{semantic}

@@

\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}

@@

LISP is the best example of a language that unifies many language constructs using a single
form. In LISP, everything is an s-experssion. That is, either a list or a symbol. In contrast, 
a typical data processing language uses a number of distinct constructs including indexers (for 
range selection and filtering), method calls (for transformations) and named parameters (for further 
configuration). Consider filtering and sorting:
%
\begin{equation*}
\begin{array}{l}
\ident{data}[\ident{data}[\str{Games}]==\str{Rio (2016)}] \qquad\ball{1}\\
\ident{data}.\ident{filter}(\kvd{fun}~\ident{row}\rightarrow\ident{row}.\ident{Games}==\str{Rio (2016)})\qquad\ball{2}\\
\ident{data}.\ident{sort\_values}(\ident{by}=\str{Gold}, \ident{ascending}=\kvd{False})\qquad\ball{3}\\
\end{array}
\end{equation*}

\noindent
Python and pandas uses indexers for filtering \ball{1} which can alternatively be written (e.g.~in 
LINQ) using a method taking a predicate as a lambda function \ball{2}. Operations that are 
parameterized only by column name, such as sorting in pandas \ball{3} are often methods with 
named parameters.

We aim to unify the above examples using a single language construct that offers a high-level programming
model\footnote{In contrast, s-expressions in LISP are typically used as a lower-level encoding of 
higher-level constructs. For example \ident{lambda} symbol encodes a lambda function as an s-expression.}
and can be supported by modern tooling (as discussed in Section~\ref{sec:analysis-auto}).
Member access provides an extremely simple programming construct that is, in conjunction with
the type provider mechanism capable of expressing the above data transformations in a uniform way: 
%
\begin{equation*}
\begin{array}{l}
\ident{data}.\qident{sort data}.\qident{by Gold descending}\qquad\ball{1}\\
\ident{data}.\qident{filter data}.\qident{Games is}.\qident{Rio (2016)}\qquad\ball{2}\\
\end{array}
\end{equation*}

\noindent
The member names tend to be longer and descriptive and we write them using \qident{...}.
The names are not usually typed by the user (see~Section~\ref{sec:analysis-auto}) and so the
length is not an issue when writing code. The above two examples illustrate two interesting
aspects of our approach.

\subparagraph{Members, type providers, discoverability.}
When sorting \ball{1} the member that specifies how sorting is done includes the name of the
column. This is possible because the pivot type provider tracks the column names (see 
Section~\ref{sec:analysis-col}) and provides members based on the available columns suitable
for use as sort keys. When filtering \ball{2}, the member \qident{Rio (2016)} is provided based
on the values in the data source (we discuss this furter in Section~\ref{sec:analysis-auto}).

These two examples illustrate that member access can be expressive, but it requires huge number
of types with huge number of members. Type providers address this by integration with the type
system (formalized in Section~\ref{sec:foo}) that discovers members lazily. This is why 
approaches based on code generation or pre-processors would not be viable.

Using descriptive member names is only viable when the names are discoverable. The above
code could be executed in a dynamically-typed languge that allows custom message-not-understood
handlers, but it would be impossible to get the name right when writing it. Our approach 
relies on discovering names through auto-completion as discussed in Section~\ref{sec:analysis-auto}.

\subparagraph{Are members expressive enough?}

no... but it works fine! See Section~\ref{sec:impl}.
   there are ways to make it more expressive too.

\begin{figure}
\begin{center}
\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{filter.png} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Auto-completion offering the available values of the athlete name column}
\label{fig:subtyping-diagram}
\end{figure}

@@

Autocompletion - see figure.

\subparagraph{Abstraction and data.}

- Needs concrete data set or a sample, but that's fine

\subparagraph{Values vs. types.}

- provide values? Could be $\qident{Games is}(\str{Rio (2016)})$

\subsection{Column names}
\label{sec:analysis-col}

@@

  ○ We use type providers to provide available options
  (e.g. something)

Considerations

 - renaming columns

\subsection{Dictionary syntax}
\label{sec:analysis-and}
  ○ We create the "and blah. and blah. then" pattern
  (e.g. grouping and sorting)

@@

Rio case study
